Supplementary Content

### 2022-01-25 experiments/2_supplement/html_supplement.Rmd
cleaning up, getting ready!

@@

The method for creating among strain diversity is explained in the main report. Here we show three visualisations, one for nine strains, one for three, and one for two.

@@

@@

## Replication method, no diversity

@@

## Temporal method, no diversity

@@

## Both methods overlaid, no diversity

@@

## Effect of diversity, replication method

@@

@@

@@

>>>>

# Creating among strain diversity

Here we provide additional description of the method used to create strain variation within functional groups. The following should be read only after the relevant sections of the main text.

Take for example the cyanobacteria functional group, a reference maximum growth rate parameter $G_{max,CB}$ and a reference tolerance parameter $h_{SR,CB}$. Recall that amount of among strain variation is controlled by the two meta-parameters $Div_{G_{max,CB}}$ and $Div_{h_{SR,CB}}$. To create a tradeoff, the two meta-parameters were always of different sign. The among strain variation was calculated such that the growth rate of strain $i = 1$ was a factor $1/({2Div_{G_{max,CB}}})$ of the reference growth rate, and the growth rate of the $i = n$ strain was a factor $2Div_{G_{max,CB}}$ of the reference growth rate. The growth rate of the $i = {2,...n-1}$ other strains was the reference growth rate multiplied by a factor that was equally distributed between the factor of the $i = 1$ and $i = n$ strain. This implementation of the variation and trade-off has the assumption of a linear trade-off of log-log transformed among-strain variation. Hence, for example, with the reference growth rate $G_{max,CB} = 0.05$, among strain variation of $Div_{G_{max,CB}} = 1$, and nine strains ($n = 9$), the growth rates of the nine strains are 0.025, 0.03, 0.035, 0.042, 0.05, 0.059, 0.071, 0.084, 0.1. Put another way, the growth rate parameter of the $i$-th of the $n$ strains in a functional group was calculated as $G_{max,i} = D_i * G_{max}$ where $D_i$ is the $i$-th element of the set $D = {2f(x)Div_{G_{max}} |x ∈ N, x ≤ n}$, where $f(x) = (x − (n + 1)/2)/((n − 1)/2)$.

This procedure ensures that the range of trait values is independent of the number of strains. When there were only two strains, the two strains took the minimum and maximum traits values, as defined by the $Div$ variable for that trait. When there were three strains, one had the minimum trait value, one the maximum, and one the mean. With nine strains, one had the minimum trait value, one the maximum, one the mean, and three were equally spaced (on log scale) between the minimum and the mean, and the remaining three between the maximum and the mean.

@@

Here we show three visualisations, one for nine strains, one for three, and one for two.

@@

## Method for finding stable states

@@

### Replication method, no diversity

@@

### Temporal method, no diversity

@@

### Both methods overlaid, no diversity

@@

### Effect of diversity, replication method

@@

### Further information

Here we give some further thoughts regarding the comparison of the temporal and replication method for stable state finding.

A particularly notable difference between the two approaches is in the initial condition. In the temporal approach, the initial condition at a new level of oxygen diffusivity is the stable state at the previous level of oxygen diffusivity, while in the approach of Bush et al (2017) the initial conditions are fixed and arbitrary (though this approach is still useful for illustrating the bistability of the system). This difference in initial conditions increases the region of bistability in the temporal approach because as a tipping point approaches, the functional group that will dominate after the regime shift starts from much lower abundances than in the approach by Bush et al (2017) and thus requires stronger amelioration of the environment to take over. Nevertheless, our main conclusions about the effects of functional trait diversity are the same using either approach.

The temporal approach is particularly useful when investigating the effects of strain richness. In preliminary simulations using the Bush et al (2017) approach, we found the counterintuitive result that higher strain richness led to lower resilience. We then recalled that we chose to divide the initial total abundance of the cyanobacterial (CB) functional group equally among the strains (termed a substitutive design in biodiversity manipulation experiments). This meant that the most tolerant CB strain in a three-strain simulation had higher initial abundance than the same most tolerant CB strain in a nine-strain simulation. The greater abundance of the most CB tolerant strain made it possible for a three-strain system to develop to be oxic, whereas the nine-strain system developed to be anoxic, despite identical range of trait variation. We could have chosen to do otherwise, for example start with the least tolerant strains of cyanobacteria as more abundant when the system started with high cyanobacterial abundance, and the most tolerant strain as more abundant when the system starts with low cyanobacterial abundance. Using the temporal approach removed the need to specify initial abundances (except the very first one), and therefore results were not so determined by the assumptions used to determine the initial conditions.

## Effect of immigration / minimum abundance event

The simulations include two methods to prevent very low densities. One is to make a floor density; if densities go below this, then they are set to that floor value. Another is to regularly add a density of 1 to all biological state variables. Here are the results for adding 1.

```{r}
## Figure 5 ----
## Region of bistability by diversity and number of strains.

# ss_2s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_SS_data_1e6.RDS"))
# stab_2s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_stab_data_1e6.RDS")) %>%
#   mutate(num_strains = 2)
# ss_2s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_SS_data_1e6_sub1.RDS"))
# stab_2s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_stab_data_1e6_sub1.RDS")) %>%
#   mutate(num_strains = 2)
# 
# 
# ss_3s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_SS_data_1e6.RDS"))
# stab_3s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_stab_data_1e6.RDS")) %>%
#   mutate(num_strains = 3) 
# ss_3s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_SS_data_1e6_sub1.RDS"))
# stab_3s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_stab_data_1e6_sub1.RDS")) %>%
#   mutate(num_strains = 3)
# 
# 
# 
# 
# ss_6s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_SS_data_1e6.RDS"))
# stab_6s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_stab_data_1e6.RDS")) %>%
#   mutate(num_strains = 6) 
# ss_6s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_SS_data_1e6_sub1.RDS"))
# stab_6s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_stab_data_1e6_sub1.RDS")) %>%
#   mutate(num_strains = 6)
#

# expt_res <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_SS_data_1e6_eventdef2.RDS"))
# num_cores <- 8
# expt_res <- expt_res %>%
#   mutate(ssfind_result = list(get_total_bio(ssfind_result)))
# cluster1 <- multidplyr::new_cluster(num_cores)
# multidplyr::cluster_library(cluster1, c("microxanox", "dplyr"))
# system.time({
#   stab_data <-  expt_res %>%
#     multidplyr::partition(cluster1) %>%
#     mutate(stability_measures = list(get_stability_measures_new(ssfind_result))) %>%
#     collect() %>%
#     unnest(cols = c(stability_measures)) 
# })
# saveRDS(stab_data, here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_eventdef2.RDS"))

stab_9s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_eventdef2.RDS")) %>%
  mutate(num_strains = 9)

stab_9s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_eventdef2.RDS")) %>%
  mutate(num_strains = 9)

#ss_9s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_SS_data_1e6_sub1.RDS"))
#stab_9s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_sub1.RDS")) %>%
#  mutate(num_strains = 9)

#stab_9s_diff3 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_diff3.RDS")) %>%
#  mutate(num_strains = 9)

# all_stab <- stab_2s %>%
#   bind_rows(stab_2s_sub1) %>%
#   bind_rows(stab_3s) %>%
#   bind_rows(stab_3s_sub1) %>%
#   bind_rows(stab_6s) %>%
#   bind_rows(stab_6s_sub1) %>%
#   bind_rows(stab_9s) %>%
#   bind_rows(stab_9s_sub1) %>%
#   bind_rows(stab_9s_diff3)

all_stab <- stab_9s

CB_vars <- unique(all_stab$CB_var_gmax_s)
SB_vars <- unique(all_stab$SB_var_gmax_s)
PB_vars <- unique(all_stab$PB_var_gmax_s)

CB_all_stab <- all_stab %>%
  filter(SB_var_gmax_s == 0,
         PB_var_gmax_s == 0) %>%
  ungroup() %>%
  mutate(var_treat = "CB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))
SB_all_stab <- all_stab %>%
  filter(CB_var_gmax_s == 0,
         PB_var_gmax_s == 0)  %>%
  ungroup() %>%
  mutate(var_treat = "SB",
         var_gmax = SB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))
PB_all_stab <- all_stab %>%
  filter(CB_var_gmax_s == 0,
         SB_var_gmax_s == 0) %>%
  ungroup()  %>%
  mutate(var_treat = "PB",
         var_gmax = PB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))

for_join <- tibble(CB_var_gmax_s = sort(CB_vars),
                   SB_var_gmax_s = sort(SB_vars))
CBSB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "CB-SB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax)) %>%
  filter(PB_var_gmax_s == 0)

for_join <- tibble(CB_var_gmax_s = sort(CB_vars),
                   PB_var_gmax_s = sort(PB_vars))
CBPB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "CB-PB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax)) %>%
  filter(SB_var_gmax_s == 0)

for_join <- tibble(SB_var_gmax_s = sort(SB_vars),
                   PB_var_gmax_s = sort(PB_vars))
SBPB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "SB-PB",
         var_gmax = SB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax)) %>%
  filter(CB_var_gmax_s == 0)

for_join <- tibble(CB_var_gmax_s = sort(CB_vars),
                   SB_var_gmax_s = sort(SB_vars),
                   PB_var_gmax_s = sort(PB_vars))
CBSBPB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "CB-SB-PB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))

all_stab_results <- CB_all_stab %>%
  bind_rows(SB_all_stab) %>%
  bind_rows(PB_all_stab) %>%
  bind_rows(CBSB_all_stab) %>%
  bind_rows(CBPB_all_stab) %>%
  bind_rows(SBPB_all_stab) %>%
  bind_rows(CBSBPB_all_stab)

all_stab_results <- all_stab_results %>%
  mutate(var_treat = forcats::fct_relevel(var_treat, levels = c("CB",
                                                                "SB",
                                                                "PB",
                                                                "CB-SB",
                                                                "CB-PB",
                                                                "SB-PB",
                                                                "CB-SB-PB")))

all_stab_results_small <- select(all_stab_results, -7:-12)

#saveRDS(all_stab_results_small, here("experiments/0_ss_finding/temporal_method/data/all_stab_results_small.RDS"))

all_stab_results %>%
  #filter(var_treat == "CB") %>%
  filter(num_strains == 9) %>%
  filter(Species == "SB_tot") %>%
  ggplot(aes(x = stand_var,
             #col = as.factor(num_strains)
             )) +
  geom_line(aes(y = hyst_min_log), lwd = 0.5, alpha = 0.8) +
  geom_line(aes(y = hyst_max_log), lwd = 0.5, alpha = 0.8) +
  geom_ribbon(aes(ymin = hyst_min_log, ymax = hyst_max_log), col = "gray", alpha = 0.2) +
  facet_wrap( ~ var_treat, scales = "free_y", nrow = 3) +
  xlab("Standardised amount of trait variation\n[see text for units]") +
  ylab("Oxygen diffusivity\n[log10 uM per hour]") +
  labs(fill = "Variation in\nonly these\nfunctional groups") +
  coord_flip() +
  guides(col = guide_legend(title="Number of strains")) +
  theme_bw() +
  theme(
    legend.position="top",
    panel.background = element_blank(),
    #strip.text.x = element_blank()
  ) +
  geom_hline(yintercept = c(-8, 0), col = "grey", lwd = 3) +
  ggtitle("Bistability region for total SB density")

#ggsave(here("experiments/1_figures_main_ms/figure_5_v2_9strain_only.pdf"),
#       width = 10, height = 8)

@@

### A case where 1'000'000 hours is not long enough

@@

### Shorter simulation times

Here 1e4 hours at each oxygen diffusivity level:

```{r}
## Figure 5 ----
## Region of bistability by diversity and number of strains.

# ss_2s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_SS_data_1e6.RDS"))
# stab_2s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_stab_data_1e6.RDS")) %>%
#   mutate(num_strains = 2)
# ss_2s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_SS_data_1e6_sub1.RDS"))
# stab_2s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/2_strain_stab_data_1e6_sub1.RDS")) %>%
#   mutate(num_strains = 2)
# 
# 
# ss_3s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_SS_data_1e6.RDS"))
# stab_3s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_stab_data_1e6.RDS")) %>%
#   mutate(num_strains = 3) 
# ss_3s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_SS_data_1e6_sub1.RDS"))
# stab_3s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/3_strain_stab_data_1e6_sub1.RDS")) %>%
#   mutate(num_strains = 3)
# 
# 
# 
# 
# ss_6s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_SS_data_1e6.RDS"))
# stab_6s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_stab_data_1e6.RDS")) %>%
#   mutate(num_strains = 6) 
# ss_6s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_SS_data_1e6_sub1.RDS"))
# stab_6s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/6_strain_stab_data_1e6_sub1.RDS")) %>%
#   mutate(num_strains = 6)
#

# expt_res <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_SS_data_1e6_eventdef2.RDS"))
# num_cores <- 8
# expt_res <- expt_res %>%
#   mutate(ssfind_result = list(get_total_bio(ssfind_result)))
# cluster1 <- multidplyr::new_cluster(num_cores)
# multidplyr::cluster_library(cluster1, c("microxanox", "dplyr"))
# system.time({
#   stab_data <-  expt_res %>%
#     multidplyr::partition(cluster1) %>%
#     mutate(stability_measures = list(get_stability_measures_new(ssfind_result))) %>%
#     collect() %>%
#     unnest(cols = c(stability_measures)) 
# })
# saveRDS(stab_data, here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_eventdef2.RDS"))

#stab_9s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_eventdef2.RDS")) %>%
#  mutate(num_strains = 9)

stab_9s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e4_eventdef2.RDS")) %>%
  mutate(num_strains = 9)

#ss_9s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_SS_data_1e6_sub1.RDS"))
#stab_9s_sub1 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_sub1.RDS")) %>%
#  mutate(num_strains = 9)

#stab_9s_diff3 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_stab_data_1e6_diff3.RDS")) %>%
#  mutate(num_strains = 9)

# all_stab <- stab_2s %>%
#   bind_rows(stab_2s_sub1) %>%
#   bind_rows(stab_3s) %>%
#   bind_rows(stab_3s_sub1) %>%
#   bind_rows(stab_6s) %>%
#   bind_rows(stab_6s_sub1) %>%
#   bind_rows(stab_9s) %>%
#   bind_rows(stab_9s_sub1) %>%
#   bind_rows(stab_9s_diff3)

all_stab <- stab_9s

CB_vars <- unique(all_stab$CB_var_gmax_s)
SB_vars <- unique(all_stab$SB_var_gmax_s)
PB_vars <- unique(all_stab$PB_var_gmax_s)

CB_all_stab <- all_stab %>%
  filter(SB_var_gmax_s == 0,
         PB_var_gmax_s == 0) %>%
  ungroup() %>%
  mutate(var_treat = "CB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))
SB_all_stab <- all_stab %>%
  filter(CB_var_gmax_s == 0,
         PB_var_gmax_s == 0)  %>%
  ungroup() %>%
  mutate(var_treat = "SB",
         var_gmax = SB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))
PB_all_stab <- all_stab %>%
  filter(CB_var_gmax_s == 0,
         SB_var_gmax_s == 0) %>%
  ungroup()  %>%
  mutate(var_treat = "PB",
         var_gmax = PB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))

for_join <- tibble(CB_var_gmax_s = sort(CB_vars),
                   SB_var_gmax_s = sort(SB_vars))
CBSB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "CB-SB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax)) %>%
  filter(PB_var_gmax_s == 0)

for_join <- tibble(CB_var_gmax_s = sort(CB_vars),
                   PB_var_gmax_s = sort(PB_vars))
CBPB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "CB-PB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax)) %>%
  filter(SB_var_gmax_s == 0)

for_join <- tibble(SB_var_gmax_s = sort(SB_vars),
                   PB_var_gmax_s = sort(PB_vars))
SBPB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "SB-PB",
         var_gmax = SB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax)) %>%
  filter(CB_var_gmax_s == 0)

for_join <- tibble(CB_var_gmax_s = sort(CB_vars),
                   SB_var_gmax_s = sort(SB_vars),
                   PB_var_gmax_s = sort(PB_vars))
CBSBPB_all_stab <- all_stab %>%
  right_join(for_join)  %>%
  ungroup() %>%
  mutate(var_treat = "CB-SB-PB",
         var_gmax = CB_var_gmax_s,
         stand_var = var_gmax / max(var_gmax))

all_stab_results <- CB_all_stab %>%
  bind_rows(SB_all_stab) %>%
  bind_rows(PB_all_stab) %>%
  bind_rows(CBSB_all_stab) %>%
  bind_rows(CBPB_all_stab) %>%
  bind_rows(SBPB_all_stab) %>%
  bind_rows(CBSBPB_all_stab)

all_stab_results <- all_stab_results %>%
  mutate(var_treat = forcats::fct_relevel(var_treat, levels = c("CB",
                                                                "SB",
                                                                "PB",
                                                                "CB-SB",
                                                                "CB-PB",
                                                                "SB-PB",
                                                                "CB-SB-PB")))

all_stab_results_small <- select(all_stab_results, -7:-12)

#saveRDS(all_stab_results_small, here("experiments/0_ss_finding/temporal_method/data/all_stab_results_small.RDS"))

all_stab_results %>%
  #filter(var_treat == "CB") %>%
  filter(num_strains == 9) %>%
  filter(Species == "SB_tot") %>%
  ggplot(aes(x = stand_var,
             #col = as.factor(num_strains)
             )) +
  geom_line(aes(y = hyst_min_log), lwd = 0.5, alpha = 0.8) +
  geom_line(aes(y = hyst_max_log), lwd = 0.5, alpha = 0.8) +
  geom_ribbon(aes(ymin = hyst_min_log, ymax = hyst_max_log), col = "gray", alpha = 0.2) +
  facet_wrap( ~ var_treat, scales = "free_y", nrow = 3) +
  xlab("Standardised amount of trait variation\n[see text for units]") +
  ylab("Oxygen diffusivity\n[log10 uM per hour]") +
  labs(fill = "Variation in\nonly these\nfunctional groups") +
  coord_flip() +
  guides(col = guide_legend(title="Number of strains")) +
  theme_bw() +
  theme(
    legend.position="top",
    panel.background = element_blank(),
    #strip.text.x = element_blank()
  ) +
  geom_hline(yintercept = c(-8, 0), col = "grey", lwd = 3) +
  ggtitle("Bistability region for total SB density")

#ggsave(here("experiments/1_figures_main_ms/figure_5_v2_9strain_only.pdf"),
#       width = 10, height = 8)
